---
title: "Gene regulatory network inference"
author: 
- name: Fabricio Almeida-Silva
  affiliation: Universidade Estadual do Norte Fluminense Darcy Ribeiro, RJ, Brazil
- name: Thiago Motta Venancio
  affiliation: Universidade Estadual do Norte Fluminense Darcy Ribeiro, RJ, Brazil
output: 
  BiocStyle::html_document:
    toc: true
    highlight: pygments
    number_sections: yes
bibliography: vignette2.bib
vignette: >
  %\VignetteIndexEntry{Gene regulatory network inference with BioNERO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = FALSE,
  cache = FALSE,
  fig.align = 'center',
  fig.width = 5,
  fig.height = 4
)
```

# Installation

```{r installation, eval=FALSE}
if(!requireNamespace('BiocManager', quietly = TRUE))
  install.packages('BiocManager')

BiocManager::install("BioNERO")
```

```{r load_package}
# Load package after installation
library(BioNERO)
set.seed(123) # for reproducibility
```


# Introduction and algorithm description

In the previous vignette, we explored all aspects of gene coexpression networks (GCNs), which are represented as **undirected weighted graphs**. It is **undirected** because, for a given link between *gene A* and *gene B*, we can only say that these genes are coexpressed, but we cannot know whether *gene A* controls *gene B* or otherwise. Further, **weighted** means that some coexpression relationships between gene pairs are stronger than others. In this vignette, we will demonstrate how to infer gene regulatory networks (GRNs) from expression data with BioNERO. GRNs display interactions between regulators (e.g., transcription factors or miRNAs) and their targets (e.g., genes). Hence, they are represented as **directed unweighted graphs**.

Numerous algorithms have been developed to infer GRNs from expression data. However, the algorithm performances are highly dependent on the benchmark data set. To solve this uncertainty, @Marbach2012 proposed the application of the *"wisdom of the crowds"* principle to GRN inference. This approach consists in inferring GRNs with different algorithms, ranking the interactions identified by each method, and calculating the average rank for each interaction across all algorithms used. This way, we can have consensus, high-confidence edges to be used in biological interpretations. For that, `BioNERO` implements three popular algorithms: GENIE3 [@Huynh-Thu2010], ARACNE [@Margolin2006] and CLR [@Faith2007].


# Data preprocessing

Before inferring the GRN, we will preprocess the expression data the same way we did in the previous vignette.

```{r}
# Load example data set
data(zma.se)

# Preprocess the expression data
final_exp <- exp_preprocess(zma.se, min_exp = 10, variance_filter = TRUE, n=2000)
```


# Gene regulatory network inference

`BioNERO` requires only 2 objects for GRN inference: the **expression data** (SummarizedExperiment, matrix or data frame) and a character vector of **regulators** (transcription factors or miRNAs). The transcription factors used in this vignette were downloaded from PlantTFDB 4.0 [@Jin2017].

```{r load_tfs}
data(zma.tfs)
head(zma.tfs)
```

## Consensus GRN inference

Inferring GRNs based on the *wisdom of the crowds* principle can be done with a single function: `exp2grn()`. This function will infer GRNs with GENIE3, ARACNE and CLR, calculate average ranks for each interaction and filter the resulting network based on the optimal scale-free topology (SFT) fit. In the filtering step, *n* different networks are created by subsetting the top *n* quantiles. For instance, if a network of 10,000 edges is given as input with `nsplit = 10`, 10 different networks will be created: the first with 1,000 edges, the second with 2,000 edges, and so on, with the last network being the original input network. Then, for each network, the function will calculate the SFT fit and select the best fit.

```{r exp2grn, fig.small=TRUE}
# Using 10 trees for demonstration purposes. Use the default: 1000
grn <- exp2grn(exp = final_exp, regulators = zma.tfs$Gene, nTrees=10)
head(grn)
```


## Algorithm-specific GRN inference

This section is directed to users who, for some reason (e.g., comparison, exploration), want to infer GRNs with particular algorithms. The available algorithms are:

**GENIE3:** a regression-tree based algorithm that decomposes the prediction of GRNs for *n* genes into *n* regression problems. For each regression problem, the expression profile of a target gene is predicted from the expression profiles of all other genes using random forests (default) or extra-trees.

```{r genie3}
# Using 10 trees for demonstration purposes. Use the default: 1000
genie3 <- grn_genie3(final_exp, regulators = zma.tfs$Gene, nTrees=10)
head(genie3)
dim(genie3)
```


**ARACNE:** information-theoretic algorithm that aims to remove indirect interactions inferred by coexpression.

```{r aracne}
aracne <- grn_aracne(final_exp, regulators = zma.tfs$Gene)
head(aracne)
dim(aracne)
```


**CLR:** extension of the relevance networks algorithm that uses mutual information to identify regulatory interactions.

```{r clr}
clr <- grn_clr(final_exp, regulators = zma.tfs$Gene)
head(clr)
dim(clr)
```


Users can also infer GRNs with the 3 algorithms at once using the function `exp_combined()`. The resulting edge lists are stored in a list of 3 elements. [^1]

[^1]: **NOTE:** Under the hood, `exp2grn()` uses `exp_combined()` followed by averaging ranks with `grn_average_rank()` and filtering with `grn_filter()`.

```{r grn_combined}
grn_list <- grn_combined(final_exp, regulators = zma.tfs$Gene, nTrees=10)
head(grn_list$genie3)
head(grn_list$aracne)
head(grn_list$clr)
```

# Gene regulatory network analysis

After inferring the GRN, `BioNERO` allows users to perform some common downstream analyses.

## Hub gene identification

GRN hubs are defined as the top 10% most highly connected regulators, but this percentile is flexible in `BioNERO`.[^2] They can be identified with `get_hubs_grn()`. 

[^2]: **NOTE:** Remember: GRNs are represented as **directed** graphs. This implies that only regulators are taken into account when identifying hubs. The goal here is to identify regulators (e.g., transcription factors) that control the expression of several genes.

```{r get_hubs}
hubs <- get_hubs_grn(grn)
hubs
```

## Network visualization


```{r plot_static, fig.height=4, fig.width=4}
plot_grn(grn)
```

GRNs can also be visualized interactively for exploratory purposes.

```{r plot_interactive}
plot_grn(grn, interactive = TRUE, dim_interactive = c(500,500))
```

Finally, `BioNERO` can also be used for visualization and hub identification in protein-protein (PPI) interaction networks. The functions `get_hubs_ppi()` and `plot_ppi()` work the same way as their equivalents for GRNs (`get_hubs_grn()` and `plot_grn()`).


# Session information {.unnumbered}

This vignette was created under the following conditions:

```{r}
sessionInfo()
```

# References
